Bayesian Inference in Numerical Cognition: A Tutorial Using JASP

### 2019-12-09 jncTutorial/manuscript.Rmd
Linear regression theory

Like any other discipline within the psychological and behavioral sciences, researchers in numerical cognition rely on statistical inference, and hypothesis testing in particular, to verify theoretical claims against observed data. In recent years, Bayesian inference has become increasingly popular as an alternative to classical null hypothesis significance testing (NHST). Bayesian inference operates under a fundamentally different framework from NHST; instead of hard "accept/reject" decisions about a null hypothesis, Bayesian inference works by quantifying knowledge about competing hypotheses in light of observed data. Hypotheses that do a better job of predicting data will receive increased support, whereas hypotheses that don't predict the data well will receive decreased support.

@@

In this tutorial paper, we will introduce the reader to the basics of Bayesian inference through the lens of some classic, well-cited studies in numerical cognition. In three detailed examples illustrating \( t \)-tests, linear regression, and ANOVA, we will walk the reader through performing and interpreting the results of Bayesian analyses using the software package JASP [@jaspSoftware]. JASP is a free, open-source statistical software package offering the user the option of both Bayesian and frequentist analyses, all housed within a familiar, easy-to-use graphical user interface. Below, we will give a brief introduction to Bayesian hypothesis testing, followed by the three tutorial examples.

# Introduction to Bayesian hypothesis testing

<!-- To elaborate on the concept of (Bayesian) hypothesis testing,  -->

In this section, we will review and elaborate on the concept of hypothesis testing. Even though the reader of this paper is probably familiar with the general concept of hypothesis testing, we believe that it is nonetheless instructive to go over the basics.

## Hypotheses
Hypothesis testing can be viewed as a process for validating general claims based on a small sample of the general population. To illustrate, suppose that we are policy makers confronted with a new study programme which claims to increase the mathematical abilities of sixth graders in the United States. To ease the exposition we use a (fictitious) scale called the Scale for Advancing Mathematical Ability (SAMA) to measure mathematical ability. As policy makers we know that the population mean SAMA score is \( \mu_{0}=50 \) and that the standard deviation is \( \sigma=10 \) for the population of sixth graders that are taught using the standard materials in the US.

The superiority claim about the new study programme can be formalised as the working hypothesis \( \mathcal{H}_{1} : \mu > 50 \). Here \( \mu \) refers to the \emph{population} mean of SAMA score of the same sixth graders, but taught using the new study programme instead. Such a treatment population does not (yet) exist and \( \mu \) cannot be measured, and is, therefore, unknown.

Nonetheless, we can infer the validity of the working hypothesis \( \mathcal{H}_{1} \) by running an experiment in which \( n=90 \) sixth graders are taught with the new study programme. Suppose that these particular participants scored an average SAMA score of \( \bar{x} = 52.07 \). Clearly, this sample mean is larger than the status quo population mean of \( 50 \), but can we assume that this finding generalises to the whole population? If the answer is "yes" and we declare the working hypothesis to be verified, then how confident would we be about such a decision? Below we recall how \( p \)-values are typically used to make hard "yes" and "no" decisions, and we remind the reader why a \( p \)-value does not quantify the (un)certainty of such decisions.

## Statistical models
To deduce how hypotheses, i.e., claims about the population such as \( \mu \), affect the data, such as \( \bar{x} \), we have to link the former to the latter with a distribution. For instance, let us suppose a normal distribution provides this link. The population parameters, the distribution that links them to the data, and the data together make up a (statistical) model, which allows for quantitative predictions. For instance, the claim that the treatment population is identitical to the current population of sixth graders implies the null model \( \mathcal{M}_{0} \) comprised of (1) the null hypothesis \( \mathcal{H}_{0}: \mu = 50 \), (2) the assumption \( \sigma = 10 \), and (3) normally distributed SAMA scores. When all three assumptions hold true, we can mathematically deduce that the \emph{potential} average SAMA score consisting of a group of \( n = 90 \) participants is expected to fall within the interval \( 47.93 = 50 - 1.96 \frac{10}{\sqrt{n}} \) and \( 52.07 = 50 + 1.96 \frac{10}{\sqrt{n}} \) with 95% chance. <!-- 
-->
\footnote{We use ``chance'' to refer to probabilistic statements about the data.}
<!-- 
-->

@@

Under the null model \( \mathcal{M}_{0} \), the more extreme data are potential outcomes of \( \bar{x} \) larger than the observed sample mean of \( 52.07 \), but also lower than the unobserved possible sample mean of \( 47.93 = 50 - (52.07-50) \). Under the null, and the normality assumptions in particular, we can deduce that the chance of seeing a potential \( \bar{x} \) in \( (- \infty, 47.93] \) or \( [52.07, \infty) \) is 5%, thus, \( p=0.05 \). Note that \( 1 - p = 0.95 \), which is the 95% chance of the non-extreme event of having \( \bar{x} \) fall within the interval \( [47.93, 52.07] \), if the null model holds true. We use this example to emphasise that \( p \)-values help us reason from the null model to (potential) data, but they do not allow us to make quantitative statements from the actually observed data about the hypotheses.

@@

<!--
the \( p \)-value does not provide us with a measure of uncertainty about the hypotheses, because they are chance statements about potential data, and not about the hypotheses.

The latter implies that one can only come to the conclusion of accept or reject, without  -->

\subsection{Bayesian testing}
Bayesian testing addresses these problems as follows. First, it compares the two hypotheses on equal footing by also operationalising the alternative model. For the case at hand, \( \mathcal{M}_{1} \) would be comprised of the assumptions (1') \( \mathcal{H}_{1}: \mu > 50 \), (2) \( \sigma =10 \) as before, and (3) the data being normally distributed as before. By doing so, violations of normality and the assumed standard deviation will penalize both models equally. Secondly, to reason from a particular observed outcome to the general population, it employs Bayes rule. To do so, we are required to express our uncertainty about the two hypotheses in terms of so-called *prior model probabilities* before data observation.

Prior model probabilities help us contextualise the testing problem. Working as policy makers, we might have seen many claims of improved study programmes that did not work out well. For instance, past claims of improved study programmes led to null results four times out of five (i.e., we place 4-to-1 odds against the claim of improvement), we might set \( P( \mathcal{M}_{1}) = 0.2 \) and \( P(\mathcal{M}_{0}) = 0.8 \). On the other hand, the designers of the new study programme to increase SAMA scores might feel strongly about the benefits of the problem, and so might choose \( P( \mathcal{M}_{1}) = 0.7 \) and \( P(\mathcal{M}_{0}) = 0.3 \) instead. If \( \mathcal{M}_{1} \) and \( \mathcal{M}_{0} \) are the only models under consideration we have \( 0 < P( \mathcal{M}_{0}) < 1 \) and \( P( \mathcal{M}_{1})= 1 - P( \mathcal{M}_{0}) \). The Bayes factor \( \textnormal{BF}_{10}(d) \) is a function of the observed data \( d \) that updates these prior model probabilities to posterior model probabilities, as follows:

@@

Note that the data only enter the formula via the Bayes factor. For instance, if \( \textnormal{BF}_{10}(d) = 6 \), then the sceptic's probability of the superiority of the new study programme increases to 60%, i.e., \( P(\mathcal{M}_{1} \mid d)=0.60 \), and retains 40% probability of for the null after data observation. On the other hand, the proponent of the new study programme would report a posterior model probability of \( P(\mathcal{M}_{1} \mid d)=0.93 \) and \( P(\mathcal{M}_{0} \mid d)=0.07 \). Note that the posterior model probabilities combine what we believed prior to data observation with the evidence of the data in terms of the Bayes factors. Both sceptic and the proponent could, if they wish, declare the new study programme to be superior to the status quo, but the sceptic would be much less confident than the proponent. The resulting posterior model probabilities can, thus, be used to complement a decision with a measure of uncertainty.

@@

Unlike the \( p \)-value, there are no universal standards for when a Bayes factor provides a sufficient amount of evidence. However, people have made recommendations in the past regarding sizes of Bayes factors, and these provide us with a helpful, but very rough guideline on its interpretation, when we do not know the context of the problem. In particular, we mention the recommendations of @jeffreys1961, who proposed the following scheme\footnote{We advise readers not to get too hung up on the specific labels and breakpoints in this table, see also @giollaInpresswhat.}:

@@

In principle, the Bayes factor does not come with a specified decision rule. We believe that context made explicit using prior model probabilities by field experts are much more important, than overly generalised "rules" such \( p < 0.05 \) or \( \textnormal{BF} _{10} > k \), where \( k \) is some arbitrarily chosen number.

# Tutorial examples
In this section, we describe a number of practical examples of using Bayesian inference in the context of numerical cognition. In the examples, each based on a well-cited paper in numerical cognition, we demonstrate a number of JASP's features using synthetic data sets, each generated in R using specific model assumptions that closely match the patterns of observed data described in the original papers.  The data sets, as well as the R script used to generate the data sets, are available for download from https://git.io/JeXui.

## Example 1: a Bayesian \( t \)-test

To demonstrate the Bayesian \( t \)-test, we will use a data set based on @vergutsDeMoor2005, who investigated whether two-digit numbers are represented holistically on a mental number line, or instead are processed by decomposing into separate digit components. In their experiment, @vergutsDeMoor2005 presented subjects with pairs of two-digit numbers who had the same decade (e.g., 53 versus 51) or decades that differed by one (e.g., 53 versus 61).  Their reasoning for this manipulation was as follows: if subjects are using holistic processes to compare two-digit numbers, then they should exhibit a numerical distance effect [@moyer1967] on same-decade as well as different-decade pairs. That is, response times should be faster when the numerical distance between the two numbers is large, and slower when the numerical distance is small.  If, on the other hand, subjects are using decomposed processes, this numerical distance effect should disappear on the different-decade pairs. This is because in different-decade pairs, such as 53 versus 61, the comparison can be made solely on the basis of the decade digit alone. Thus, response times in this condition should be immune to manipulation of numerical distance.

@@

### Theory of the Bayesian \( t \)-test

The Bayesian \( t \)-test forms the basis for the other tests described in this paper. The technical details are included for completeness, but are not necessary for the use of Bayes factors.

@@

The Bayesian \( t \)-test is based on the priors \( \pi_{1} \), which puts an "uninformative" prior on the nuisance parameter \( \sigma \) [e.g., @ly2017tutorial] in conjunction with a Cauchy prior on the population effect size \( \delta = \mu / \sigma \), and the prior \( \pi_{0} \) with the same "uninformative" prior on the nuisance parameter \( \sigma \). With these priors, the resulting Bayes factor has a number of desirable properties: (1) \textbf{Measurement invariant.} The Bayes factor as a measure of evidence does not change if the units of measurements are changed [e.g., @rouder2009bayesian]. For instance, if the data were to consist of temperature measurements, the Bayes factor would be the same regardless of the data being entered as Celsius or Fahrenheit. (2) \textbf{Predictively matched.} The Bayes factor reflects perfectly indifference, i.e., \( \textnormal{BF}_{10}=1 \) when the data are completely uninformative. A Bayes factor of one implies that the posterior model probabilities equal the prior model probabilities, and nothing has been learned; see @jeffreys1961, for further details. (3) \textbf{Information consistent.} For a sufficiently large enough sample with size \( n \) fixed, the Bayes factor provides overwhelmingly support for the alternative, whenever \( t \)-value goes to infinity; see @gronauInpressinformed, @ly2016harold for further details. (4) \textbf{(Asymptotically) consistent.} As the Bayes factor takes in an increasing number of samples of data points generated under the null model, then as \( n \) grows, the Bayes factor \( \textnormal{BF}_{10} \) tends to zero to convey support for the data generating model \( \mathcal{M}_{0} \). On the other hand, if \( n \) grows with data generated under the alternative, the Bayes factor \( \textnormal{BF}_{10} \) grows unbounded; see for instance @liang2008mixtures. (5) \textbf{Robust to changes in the sample size.} With the uninformative prior on \( \sigma \), this Bayes factor allows for optional stopping, that is, early stopping when there is enough evidence for the alternative [e.g., @hendriksen2018optional] without inflating the type 1 error. On top of this, the Bayesian \( t \)-test is also allows for optional continuation, which implies that one can build on top of previous results; see for instance @faulkenberry2019estimating, @grunwald2019safe, @ly2017replication. Note that properties (4), (5), and (6) imply that the Bayesian \( t \)-test, first derived by @jeffreys1948theory2, also has good frequentist properties.

@@

We will now describe a Bayesian analysis inspired by the original methods of @vergutsDeMoor2005. After downloading the file \texttt{ttest.csv} from the Github repository (https://git.io/JeXui), you can open the file in JASP by clicking the menu button -- "Open" -- "Computer", which will then open a file browser for you to locate the downloaded file on your computer. Once the file is opened, you will be able to see a data set with three variables (see Figure \@ref(fig:ttestData). The variable \texttt{condition} defines whether a participant was instructed to choose the smaller or the larger of each presented two-digit number pair. The variable \texttt{distSame} represents each participant's slope after regressing response times against numerical distance [c.f., @lorch1990] for same-decade pairs. Note that a negative slope is indicative of the numerical distance effect, as response times decrease as a function of increasing numerical distance. Finally, \texttt{distDifferent} represents each subject's regression slope for different-decade pairs.

@@

We are now ready to perform a Bayesian analysis of the simulated @vergutsDeMoor2005 data. First, since we will separately analyze the "Choose smaller" and "Choose larger" groups, we will need to apply a *filter* to the data set; specifically, we will first consider only the "Choose smaller" group. <!--@Tom, Would it be easier if we explained this using the click filter -->There are several ways to do this in JASP, but the easiest is the "click filter"\footnote{For a demonstration of the different methods of filtering in JASP, see https://jasp-stats.org/2018/06/27/how-to-filter-your-data-in-jasp/.} First, click the header of the \texttt{condition} column. To remove all of the \texttt{larger} trials from the analysis and retain only the \texttt{smaller} trials, we simply "turn off" the \texttt{larger} value by clicking on the check mark next to it. The checkmark will be replaced by a cross, and you will be able to see that the \texttt{larger} trials have been greyed out. The filter dialog can be closed by clicking on the X on the right side of the dialog box. Notice that a funnel icon will appear next to the \texttt{condition} column header to indicate that a filter is currently being applied. Any filters can be removed by re-opening the filter dialog and clicking the eraser icon.

To perform a Bayesian one-sample \( t \)-test, click on the "T-Tests" button in the ribbon at the top, and select "Bayesian One Sample T-Test". There are two \( t \)-tests that we wish to perform, but for clarity, we will start with describing one in detail. Before beginning, however, we define our two competing hypotheses explicitly. Since the numerical distance effect is expected to be negative, we define \( \mathcal{H}_{0}:\delta = 0 \) and \( \mathcal{H}_{-}:\delta < 0 \), where the minus in the subscript reminds of that the alternative is one-sided negative.

@@

The Bayes factor is \( BF_{-0}=9.027 \), which indicates that the observed data are approximately 9 times more likely under the alternative hypothesis \( \mathcal{H}_{1} \) than under the null hypothesis \( \mathcal{H}_{0} \). According to the convention of Jeffreys (1961), this is considered "moderate" evidence for \( \mathcal{H}_{1} \), which states that the true population effect \( \delta \) (from which our participants' regression slopes are drawn) is negative. A useful visual display of this Bayes factor is provided in the plot in the bottom of the Results pane. This probability wheel (or "pizza plot") depicts the Bayes factor as an odds ratio of 9:027-to-1 in favor of \( \mathcal{H}_{1} \) over the null.\footnote{To simplify matters, we identify the hypotheses to the models and use the notation \( \mathcal{H}_{k} \) interchangeably with \( \mathcal{M}_{k} \) for the \( k \)th model.}

The bottom plot shows this in more detail. The dashed line depicts the aforementioned Cauchy prior (with a default scale/prior width of \( r=1/\sqrt{2} = 0.707 \)) on the population effect size \( \delta \), but restricted to negative values in accordance to the alternative \( \mathcal{H}_{-} \). The solid line depicts the posterior density for \( \delta \). Under the assumption that the effect cannot be positive, this density provides a measure of uncertainty about the size of the population effect \( \delta \) *after* observing the data. We can see in Figure \@ref(fig:ttestBayes) that the more probable values of \( \delta \) are now centered around \( \delta = -0.5 \). For the testing problem, one specific value worth exploring is \( \delta=0 \). Notice that relative to the prior, the \( y \)-coordinate at \( \delta=0 \) in the posterior is smaller than the \( y \)-coordinate at \( \delta=0 \). This reflects the fact that our posterior belief that regarding \( \mathcal{H}_{0} : \delta = 0 \) has decreased after data observation. In fact, it has decreased by a factor of approximately 9, which is the Bayes factor we obtained. This relationship is no accident; For the \( t \)-test the Bayes factor can be represented as the ratio of the prior and posterior on the population effect \( \delta \) evaluated at the null hypothesis \( \mathcal{H}_{0}: \delta = 0 \). This fact is known as the Savage-Dickey density ratio [@wagenmakers2010].

@@

One last question to consider in our analysis is how sensitive the Bayes factor is to changes in the specification of the Cauchy prior on the population effect \( \delta \). Clicking the box for "Bayes factor robustness check" will help us to answer this question. Figure \@ref(fig:ttestRobust) displays the result of this robustness check. As mentioned before, the default prior for \( \delta \) is a Cauchy distribution centered around \( 0 \) with scale (or "width") \( r=0.707\). Figure \@ref(fig:ttestRobust) depicts how the Bayes factor changes with varying values of \( r \). The plot represents the Bayes factor as a continuous function of \( r \), and the legend at the top lists the Bayes factor at four values: the value of \( r \) that produces the maximum Bayes factor, the original user-specified prior (\( r=0.707 \)), a "wide" prior (\( r=1 \)), and an "ultrawide" prior (\( r=\sqrt{2}=1.414 \)). This plot should be interpreted holistically - across a wide range of values for the prior scale \( r \), the resulting Bayes factor ranges from 6.3 to 9, which implies that the result is quite robust.

@@

## Example 2: a Bayesian linear regression

>>>>

Like any other discipline within the psychological and behavioral sciences, researchers in numerical cognition rely on statistical inference, and hypothesis testing in particular, to verify theoretical claims against observed data. In recent years, Bayesian inference has become increasingly popular as an alternative to classical null hypothesis significance testing (NHST). Bayesian inference operates under a fundamentally different framework from NHST; instead of hard "accept/reject" decisions about a null hypothesis, Bayesian inference works by quantifying evidence about competing hypotheses in light of observed data. Hypotheses that do a better job of predicting data will receive increased support, whereas hypotheses that do not predict the data well will receive decreased support.

@@

In this tutorial paper, we will introduce the reader to the basics of Bayesian inference through the lens of some classic, well-cited studies in numerical cognition. In three detailed examples illustrating \( t \)-tests, linear regression, and ANOVA, we will walk the reader through performing and interpreting the results of Bayesian analyses using the software package JASP [@jaspSoftware]. JASP is a free, open-source statistical software package offering the user the option of both Bayesian and frequentist analyses, all housed within a familiar, easy-to-use graphical user interface. Below, we give a brief introduction to Bayesian hypothesis testing, followed by the three tutorial examples.

# Introduction to Bayesian hypothesis testing
In this section, we will review and elaborate on the concept of hypothesis testing. Even though the reader of this paper is probably familiar with the general concept of hypothesis testing, we believe that it is nonetheless instructive to go over the basics.

## Hypotheses
Hypothesis testing can be viewed as a process of validating general claims based on a small sample of the general population. To illustrate, suppose that we are policy makers confronted with a new study program which claims to increase the mathematical abilities of sixth graders in the United States. To ease the exposition we use a (fictitious) scale called the Scale for Advancing Mathematical Ability (SAMA) to measure mathematical ability. As policy makers we know that the population mean SAMA score is \( \mu_{0}=50 \) and that the standard deviation is \( \sigma=10 \) for the population of sixth graders that are taught using the standard materials in the US.

The superiority claim about the new study programme can be formalized as the working hypothesis \( \mathcal{H}_{1} : \mu > 50 \). Here \( \mu \) refers to the \emph{population} mean of SAMA score of the same sixth graders, but taught using the new study program instead. Such a treatment population does not (yet) exist and \( \mu \) cannot be measured, and is, therefore, unknown.

Nonetheless, we can infer the validity of the working hypothesis \( \mathcal{H}_{1} \) by running an experiment in which, say, \( n=90 \) sixth graders are taught with the new study programme. Suppose that these particular participants scored an average SAMA score of \( \bar{y} = 52.07 \). Clearly, this sample mean is larger than the status quo population mean of \( 50 \), but can we assume that this finding generalizes to the whole population? If the answer is "yes" and we declare the working hypothesis to be verified, then how confident would we be about such a decision? Below we recall how \( p \)-values are typically used to make hard "yes" and "no" decisions, and we remind the reader why a \( p \)-value does not quantify the (un)certainty of such decisions.

## Statistical models
To deduce how hypotheses, i.e., claims about the population such as \( \mu \), affect the data, such as \( \bar{y} \), we have to link the former to the latter with a distribution. For instance, let us suppose a normal distribution provides this link. The population parameters, the distribution that links them to the data, and the data together make up a (statistical) model, which allows for quantitative predictions. For instance, the claim that the treatment population is identitical to the current population of sixth graders implies the null model \( \mathcal{M}_{0} \) comprised of (1) the null hypothesis \( \mathcal{H}_{0}: \mu = 50 \), (2) the assumption \( \sigma = 10 \), and (3) normally distributed SAMA scores. When all three assumptions hold true, we can mathematically deduce that the \emph{potential} average SAMA score for a group of \( n = 90 \) participants is expected to fall within the interval \( 47.93 = 50 - 1.96 \frac{10}{\sqrt{n}} \) and \( 52.07 = 50 + 1.96 \frac{10}{\sqrt{n}} \) with 95% chance.<!-- 
-->
\footnote{We use ``chance'' to refer to probabilistic statements about the data.}
<!-- 
-->

@@

Under the null model \( \mathcal{M}_{0} \), the more extreme data are potential outcomes of \( \bar{y} \) larger than the observed sample mean of \( 52.07 \), but also lower than the unobserved possible sample mean of \( 47.93 = 50 - (52.07-50) \). Under the null, and the normality assumptions in particular, we can deduce that the chance of seeing a potential \( \bar{y} \) in \( (- \infty, 47.93] \) or \( [52.07, \infty) \) is 5%, thus, \( p=0.05 \). Note that \( 1 - p = 0.95 \), which is the 95% chance of the non-extreme event of having \( \bar{y} \) fall within the interval \( [47.93, 52.07] \), if the null model holds true. We use this example to emphasize that \( p \)-values help us reason from the null model to (potential) data, but they do not allow us to make quantitative statements from the actually observed data about the hypotheses.

@@

\subsection{Bayesian testing}
Bayesian testing addresses these problems as follows. First, it compares the two hypotheses on equal footing by also operationalizing the alternative model. For the case at hand, \( \mathcal{M}_{1} \) would be comprised of the assumptions (1') \( \mathcal{H}_{1}: \mu > 50 \), (2) \( \sigma =10 \) as before, and (3) the data being normally distributed as before. By doing so, violations of normality and the assumed standard deviation will penalize both models equally. Secondly, to reason from a particular observed outcome to the general population, it employs Bayes rule. To do so, we are required to express our uncertainty about the two hypotheses in terms of so-called *prior model probabilities* before data observation.

Prior model probabilities help us contextualize the testing problem. Working as policy makers, we might have seen many claims of improved study programs that did not work out well. For instance, if past claims of improved study programmes led to null results four times out of five (i.e., we place 4-to-1 odds against the claim of improvement), then we might set \( P( \mathcal{M}_{1}) = 0.2 \) and \( P(\mathcal{M}_{0}) = 0.8 \). On the other hand, the designers of the new study program to increase SAMA scores might feel strongly about the benefits of their  program, and so might choose \( P( \mathcal{M}_{1}) = 0.7 \) and \( P(\mathcal{M}_{0}) = 0.3 \) instead. If \( \mathcal{M}_{1} \) and \( \mathcal{M}_{0} \) are the only models under consideration we have \( 0 < P( \mathcal{M}_{0}) < 1 \) and \( P( \mathcal{M}_{1})= 1 - P( \mathcal{M}_{0}) \). The Bayes factor \( \textnormal{BF}_{10}(d) \) is a function of the observed data \( d \) that updates these prior model probabilities to posterior model probabilities, as follows:

@@

Note that the data only enter the formula via the Bayes factor. For instance, if \( \textnormal{BF}_{10}(d) = 6 \), then the sceptic's probability of the superiority of the new study program increases to 60%, i.e., \( P(\mathcal{M}_{1} \mid d)=0.60 \), and retains 40% probability of for the null after data observation. On the other hand, the proponent of the new study program would report a posterior model probability of \( P(\mathcal{M}_{1} \mid d)=0.93 \) and \( P(\mathcal{M}_{0} \mid d)=0.07 \). Note that the posterior model probabilities combine what we believed prior to data observation with the evidence of the data in terms of the Bayes factors. Both sceptic and the proponent could, if they wish, declare the new study programme to be superior to the status quo, but the sceptic would be much less confident than the proponent. The resulting posterior model probabilities can, thus, be used to complement a decision with a measure of uncertainty.

@@

Unlike the \( p \)-value, there are no universal standards for when a Bayes factor provides a sufficient amount of evidence. However, people have made recommendations in the past regarding sizes of Bayes factors, and these provide us with a helpful, but very rough guideline on its interpretation, when we do not know the context of the problem. In particular, we mention the recommendations of @jeffreys1961, who proposed the following scheme<!--
-->
\footnote{We advise readers not to get too hung up on the specific labels and breakpoints in this table, see also @giollaInpresswhat.}:

@@

In principle, the Bayes factor does not come with a specified decision rule. We believe that context made explicit using prior model probabilities by field experts are much more important, than overly generalized "rules" such \( p < 0.05 \) or \( \textnormal{BF} _{10} > k \), where \( k \) is some arbitrarily chosen number.

# Tutorial examples
In this section, we describe a number of practical examples of using Bayesian inference in the context of numerical cognition. In the examples, each based on a well-cited paper in numerical cognition, we demonstrate a number of JASP's features using synthetic data sets, each generated in R using specific model assumptions that closely match the patterns of observed data described in the original papers.The data sets, as well as the R script used to generate the data sets, are available for download from https://git.io/JeXui.

## Example 1: a Bayesian \( t \)-test
To demonstrate the Bayesian \( t \)-test, we will use a data set based on @vergutsDeMoor2005, who investigated whether two-digit numbers are represented holistically on a mental number line, or instead are processed by decomposing into separate digit components. In their experiment, @vergutsDeMoor2005 presented subjects with pairs of two-digit numbers who had the same decade (e.g., 53 versus 51) or decades that differed by one (e.g., 53 versus 61).  Their reasoning for this manipulation was as follows: if subjects are using holistic processes to compare two-digit numbers, then they should exhibit a numerical distance effect [@moyer1967] on same-decade as well as different-decade pairs. That is, response times should be faster when the numerical distance between the two numbers is large, and slower when the numerical distance is small.  If, on the other hand, subjects are using decomposed processes, this numerical distance effect should disappear on the different-decade pairs. This is because in different-decade pairs, such as 53 versus 61, the comparison can be made solely on the basis of the decade digit alone. Thus, response times in this condition should be immune to manipulation of numerical distance.

@@

### Theory of the Bayesian \( t \)-test
The Bayesian \( t \)-test forms the basis for the other tests described in this paper. The technical details are included for completeness, but are not necessary for the use of Bayes factors.

@@

The Bayesian \( t \)-test is based on the priors \( \pi_{1} \), which puts an "uninformative" prior on the nuisance parameter \( \sigma \) [e.g., @ly2017tutorial] in conjunction with a Cauchy prior on the population effect size \( \delta = \mu / \sigma \), and the prior \( \pi_{0} \) with the same "uninformative" prior on the nuisance parameter \( \sigma \). With these priors, the resulting Bayes factor has a number of desirable properties: (1) \textbf{Measurement invariant.} The Bayes factor as a measure of evidence does not change if the units of measurements are changed [e.g., @rouder2009bayesian]. For instance, if the data were to consist of temperature measurements, the Bayes factor would be the same regardless of the data being entered as Celsius or Fahrenheit. (2) \textbf{Predictively matched.} The Bayes factor reflects perfectly indifference, i.e., \( \textnormal{BF}_{10}=1 \) when the data are completely uninformative. A Bayes factor of one implies that the posterior model probabilities equal the prior model probabilities, and nothing has been learned; see @jeffreys1961, for further details. (3) \textbf{Information consistent.} For a sufficiently large enough sample size \( n \) fixed, the Bayes factor provides overwhelmingly support for the alternative, whenever the observed \( t \)-value goes to infinity; see @gronauInpressinformed, @ly2016harold for further details. (4) \textbf{(Asymptotically) consistent.} As the Bayes factor takes in an increasing number of samples of data points generated under the null model, then as \( n \) grows, the Bayes factor \( \textnormal{BF}_{10} \) tends to zero to convey support for the data generating model \( \mathcal{M}_{0} \). On the other hand, if \( n \) grows with data generated under the alternative, the Bayes factor \( \textnormal{BF}_{10} \) grows unbounded; see for instance @liang2008mixtures. (5) \textbf{Robust to changes in the sample size.} With the uninformative prior on \( \sigma \), this Bayes factor allows for optional stopping, that is, early stopping when there is enough evidence for the alternative [e.g., @hendriksen2018optional] without inflating the type 1 error. On top of this, the Bayesian \( t \)-test also allows for optional continuation, which implies that one can build on top of previous results; see for instance @faulkenberry2019estimating, @grunwald2019safe, @ly2017replication. Note that properties (4), (5), and (6) imply that the Bayesian \( t \)-test, first derived by @jeffreys1948theory2, also has good frequentist properties.

@@

### Implementation in JASP
We will now describe a Bayesian analysis inspired by the original methods of @vergutsDeMoor2005. After downloading the file \texttt{ttest.csv} from the Github repository (https://git.io/JeXui), you can open the file in JASP by clicking the menu button -- "Open" -- "Computer", which will then open a file browser for you to locate the downloaded file on your computer. Once the file is opened, you will be able to see a data set with three variables (see Figure \@ref(fig:ttestData). The variable \texttt{condition} defines whether a participant was instructed to choose the smaller or the larger of each presented two-digit number pair. The variable \texttt{distSame} represents each participant's slope after regressing response times against numerical distance [c.f., @lorch1990] for same-decade pairs. Note that a negative slope is indicative of the numerical distance effect, as response times decrease as a function of increasing numerical distance. Finally, \texttt{distDifferent} represents each subject's regression slope for different-decade pairs.

@@

We are now ready to perform a Bayesian analysis of the simulated @vergutsDeMoor2005 data. First, since we will separately analyze the "Choose smaller" and "Choose larger" groups, we will need to apply a *filter* to the data set; specifically, we will first consider only the "Choose smaller" group. There are several ways to do this in JASP, but the easiest is the "click filter"<!--
-->
\footnote{For a demonstration of the different methods of filtering in JASP, see https://jasp-stats.org/2018/06/27/how-to-filter-your-data-in-jasp/.}<!--
--> 
First, click the header of the \texttt{condition} column. To remove all of the \texttt{larger} trials from the analysis and retain only the \texttt{smaller} trials, we simply "turn off" the \texttt{larger} value by clicking on the check mark next to it. The check mark will be replaced by a cross, and you will be able to see that the \texttt{larger} trials have been greyed out. The filter dialog can be closed by clicking on the X on the right side of the dialog box. Notice that a funnel icon will appear next to the \texttt{condition} column header to indicate that a filter is currently being applied. Any filters can be removed by re-opening the filter dialog and clicking the eraser icon.

To perform a Bayesian one-sample \( t \)-test, click on the "T-Tests" button in the ribbon at the top, and select "Bayesian One Sample T-Test". There are two \( t \)-tests that we wish to perform, but for clarity, we will start with describing one in detail. Before beginning, however, we define our two competing hypotheses explicitly. Since the numerical distance effect is expected to be negative, we define \( \mathcal{H}_{0}:\delta = 0 \) and \( \mathcal{H}_{-}:\delta < 0 \), where the minus in the subscript reminds us that the alternative is one-sided negative.

@@

The Bayes factor is \( BF_{-0}=9.027 \), which indicates that the observed data are approximately 9 times more likely under the alternative hypothesis \( \mathcal{H}_{1} \) than under the null hypothesis \( \mathcal{H}_{0} \). According to the convention of Jeffreys (1961), this is considered "moderate" evidence for \( \mathcal{H}_{1} \), which states that the true population effect \( \delta \) (from which our participants' regression slopes are drawn) is negative. A useful visual display of this Bayes factor is provided in the plot in the bottom of the Results pane. This probability wheel (or "pizza plot") depicts the Bayes factor as an odds ratio of 9:027-to-1 in favor of \( \mathcal{H}_{1} \) over the null.<!--
-->
\footnote{To simplify matters, we identify the hypotheses to the models and use the notation \( \mathcal{H}_{k} \) interchangeably with \( \mathcal{M}_{k} \) for the \( k \)th model.}<!--
-->
The bottom plot shows this in more detail. The dashed line depicts the aforementioned Cauchy prior (with a default scale/prior width of \( r=1/\sqrt{2} = 0.707 \)) on the population effect size \( \delta \), but restricted to negative values in accordance to the alternative \( \mathcal{H}_{-} \). The solid line depicts the posterior density for \( \delta \). Under the assumption that the effect cannot be positive, this density provides a measure of uncertainty about the size of the population effect \( \delta \) *after* observing the data. We can see in Figure \@ref(fig:ttestBayes) that the more probable values of \( \delta \) are now centered around \( \delta = -0.5 \). For the testing problem, one specific value worth exploring is \( \delta=0 \). Notice that relative to the prior, the \( y \)-coordinate at \( \delta=0 \) in the posterior is smaller than the \( y \)-coordinate at \( \delta=0 \). This reflects the fact that our posterior belief regarding \( \mathcal{H}_{0} : \delta = 0 \) has decreased after data observation. In fact, it has decreased by a factor of approximately 9, which is the Bayes factor we obtained. This relationship is no accident; For the \( t \)-test the Bayes factor can be represented as the ratio of the prior and posterior on the population effect \( \delta \) evaluated at the null hypothesis \( \mathcal{H}_{0}: \delta = 0 \). This fact is known as the Savage-Dickey density ratio [@wagenmakers2010].

@@

One last question to consider in our analysis is how sensitive the Bayes factor is to changes in the specification of the Cauchy prior on the population effect \( \delta \). Clicking the box for "Bayes factor robustness check" will help us to answer this question. Figure \@ref(fig:ttestRobust) displays the result of this robustness check. As mentioned before, the default prior for \( \delta \) is a Cauchy distribution centered around \( 0 \) with scale (or "width") \( r=0.707\). Figure \@ref(fig:ttestRobust) depicts how the Bayes factor changes with varying values of \( r \). The plot represents the Bayes factor as a continuous function of \( r \), and the legend at the top lists the Bayes factor at four values: the value of \( r \) that produces the maximum Bayes factor, the original user-specified prior (\( r=0.707 \)), a "wide" prior (\( r=1 \)), and an "ultrawide" prior (\( r=\sqrt{2}=1.414 \)). This plot should be interpreted holistically --across a wide range of values for the prior scale \( r \), the resulting Bayes factor ranges from 6.3 to 9, which implies that the result is quite robust.

@@

## Example 2: a Bayesian linear regression
We use an example inspired by @hollowayAnsari2009, who examined the relationship between basic numerical processing and higher mathematical skills in children. Specifically, they examined whether childrens' mathematical achievement scores were related to their ability to compare symbolic and nonsymbolic number magnitudes. In this example, we will recreate one specific component of their analysis; namely, we will use Bayesian linear regression to examine the extent to which mathematical fluency scores are predicted by childrens' standardized symbolic and nonsymbolic distance effects, calculated by subtracting mean RTs for large numerical distances from mean RTs for small numerical distances and then dividing by the mean RT for large distances.<!--@Tom: The sentence above is rather hard for me to parse. Is there a possibility to split it up?-->
In their original paper, @hollowayAnsari2009 found that the childrens' symbolic distance effect was a significant predictor of fluency, \( \beta = -0.236 \), \( p<0.05 \), whereas the nonsymbolic distance effect did not predict fluency, \( \beta=0.063 \), \( p>0.05 \).
